\Title

### 2021-06-11 latex/paper/main.tex
Final edit as per meeting with Manfredi, he will now rewrite the proof

% TODO: remove onecolumn and draftcls formatting
\documentclass[journal,twoside,web, onecolumn,draftcls]{ieeecolor}

@@

%---------- Header ----------% 
\newcommand*{\Title}{VNHC and Acrobot Title}
\markboth{\journalname, VOL. XX, NO. XX, XXXX 2021}
{Moran-MacDonald \MakeLowercase{\textit{et al.}}: \Title}

@@

Our goal is to design a smooth function \(f : \Sone \times \R\) such that the
relation \(q_a = f(q_u,p_u)\) for system \eqref{eqn:acrobot-hamiltonian}
can be enforced asymptotically via feedback control (in Section \ref{sec:vnhc}
we call this a regular virtual nonholonomic constraint).  
We will further require that the dynamics of the acrobot, when the relation
holds, gain or lose energy on some set \(D\) in the sense of Definition
\ref{defn:energy-gain}.

Note that any system satisfying Definition \ref{defn:energy-gain}
can have unstable equilibria on \(D\), but not limit cycles nor closed orbits.

%/========== VNHC ==========/%
\section{Theory of VNHCs}\label{sec:vnhc}
Before embarking on our design problem, we must summarize the relevant
theory of virtual nonholonomic constraints for a class of mechanical systems we
call ``simply actuated hamiltonian systems". 
Note that the results we provide in Section \ref{sec:vnhc-vnhc} are not novel:
they are a special case of the results in
\cite{hybrid_zero_dynamics_bipedal_nhvcs}.

@@

\begin{assm}\label{assm:B-const}
    The input matrix \(B(q) \equiv B \in \R^{n\times k}\) is constant,
    full rank, and \(k < n\).
\end{assm}

\begin{assm}\label{assm:B-perp}
    There exists a right semi-orthogonal matrix 
    \(B^\perp \in \R^{(n-k)\times n}\)
    which is a left-annihilator for \(B\). 
\end{assm}

Note that Assumption \ref{assm:B-perp} requires the rows of \(B^\perp\) be unit vectors
that are mutually orthogonal. 
When \(k = (n-1)\), Assumption \ref{assm:B-perp} can be removed because it is
automatically implied by Assumption \ref{assm:B-const}.

The above assumptions allow us to define a
canonical coordinate transformation of \eqref{eqn:hamiltonian} 
which decouples the input forces.
To define this transformation we will make use of the following lemma.

@@

In light of Lemma \ref{lemma:B-orthogonal} there is no loss of generality in
assuming that the input matrix is left semi-orthogonal, which means that
\(B\tpose\) is right semi-orthogonal.
Now, let \(\mathbf{B} := [B^\perp; B\tpose]\).
Since \(B^\perp\) is a left annihilator of \(B\) and both \(B^\perp\) and
\(B\tpose\) are right semi-orthogonal, one can easily show that \(\mathbf{B}\) is
an orthogonal matrix.

We temporarily take \(\mathcal{Q}\) to be embedded in \(\R^N\) so that the
transformation \(\tilde{q} = \mathbf{B}q\) is well-defined.

\begin{thm}\label{thm:simply-actuated}
    Take the Hamiltonian system \eqref{eqn:hamiltonian} and suppose
    Assumptions \ref{assm:B-const} and \ref{assm:B-perp} hold.
    The coordinate transformation
    \(\left(\tilde{q} = \mathbf{B}q, \tilde{p} = \mathbf{B}p\right)\)
    is a canonical transformation and the resulting dynamics are given by 
    \begin{gather}\label{eqn:simple-hamiltonian}
        \mathcal{H}(\tilde{q},\tilde{p}) = 
        \frac{1}{2} \tilde{p}\tpose \Minv(\tilde{q}) \tilde{p} + V(\tilde{q})
        , \\
       \begin{cases}
           \dot{\tilde{q}} = \Minv(\tilde{q})\tilde{p}
           , \\
           \dot{\tilde{p}} = -\frac{1}{2} (\Id{n} \otimes \tilde{p}\tpose)
           \nabla_{\tilde{q}} \Minv(\tilde{q}) \tilde{p} \\
           \phantom{---} - \nabla_{\tilde{q}} V(\tilde{q}) + \simpleB \tau
            ,
        \end{cases} \nonumber
    \end{gather}
    where 
    \(\Minv(\tilde{q}) := 
    \mathbf{B}D^{-1}(\mathbf{B}\tpose \tilde{q})\mathbf{B}\tpose\)
    and
    \(V(\tilde{q}) := P(\mathbf{B}\tpose \tilde{q})\).
\end{thm}
\begin{proof}
    Since \(\mathbf{B}\) is constant, this transformation satisfies
    \(\partial\tilde{q}_i/\partial p_j = \partial\tilde{p}_i/\partial q_j = 0\) for all 
    \(i,j \in \{1,\ldots,n\}\).
    This implies the Poisson brackets \([\tilde{q}_i, \tilde{q}_j]\)
    and \([\tilde{p}_i,\tilde{p}_j]\) are both zero.
    Then, since \(\mathbf{B}\) is orthogonal, 
    \([\tilde{p}_i, \tilde{q}_j] = (\mathbf{B}_i)\tpose (\mathbf{B}\tpose)_j
        = \delta_i^j\).
    By (45.10) in \cite{landau_mechanics}, this is a canonical transformation
    and the new Hamiltonian is
    \(\mathcal{H}(\mathbf{B}\tpose \tilde{q}, \mathbf{B}\tpose \tilde{p})\).
    Finally, since \(\dot{\tilde{p}} = \mathbf{B} \dot{p}\), the input
    matrix for the system in \((\tilde{q},\tilde{p})\) coordinates is
    \(\mathbf{B}B = [\Zmat{(n-k)\times k}; \Id{k}]\), which proves the theorem.
\end{proof}

We call the \((\tilde{q},\tilde{p})\) coordinates
\textit{simply actuated coordinates}, and we call system
\eqref{eqn:simple-hamiltonian} a \textit{simply actuated hamiltonian system}.
The first \((n-k)\) configuration variables in \(\tilde{q}\), labelled \(q_u\),
are the \textit{unactuated coordinates}; 
the remaining \(k\) configuration variables, labelled \(q_a\), are the
\textit{actuated coordinates}.
The corresponding \((p_u, p_a)\) in \(\tilde{p}\) are the \textit{unactuated}
and \textit{actuated momenta}, respectively.

@@

\subsection{Energy Regulation}\label{sec:energy-reg}
One can apply the results of Theorem \ref{thm:acrobot-energy-stabilization} 
towards energy regulation;
that is, one can stabilize oscillations or rotations by appropriately toggling
between injection and dissipation VNHCs.

\textbf{Oscillation Regulation}: one chooses a desired oscillation angle 
\(q_\text{des} \in \, ]0,\pi[\) and, to avoid infinite switching, a
hysteresis value \(\delta \in [0,1]\).
Each time the orbit crosses the \(q_u\)-axis (i.e., when \(p_u = 0\)) or 
when \(|q_u| = \pi\), the supervisor does the following: 
\begin{itemize}
    \item If \(|q_u| < (1-\delta)q_\text{des}\), enable the injection VNHC.
    \item If \(|q_u| > (1+\delta)q_\text{des}\), enable the dissipation VNHC.
    \item If \((1-\delta)q_\text{des} \leq |q_u| \leq (1+\delta)q_\text{des}\),
        keep the leg extended at \(q_a = 0\). This can be done continously since
        \(q_a = 0\) when \(p_u = 0\).
\end{itemize}
Here one must be careful to choose \(\delta\) so that
\((1+\delta)q_\text{des} < \pi\), otherwise a rotating acrobot will not
dissipate energy.

\textbf{Rotation Regulation}: one chooses a desired rotation rate 
\(p_\text{des} > 0\) and a hysteresis value \(\delta \in [0,1]\).
Each time the orbit crosses the \(p_u\)-axis (i.e. when \(q_u = 0\)), the
supervisor changes which VNHC is enforced as follows:
\begin{itemize}
    \item If \(|p_u| < (1-\delta)p_\text{des}\), enable the injection VNHC.
    \item If \(|p_u| > (1+\delta)p_\text{des}\), enable the dissipation VNHC.
    \item If \((1-\delta)p_\text{des} \leq |p_u| \leq (1+\delta)p_\text{des}\),
        extend the leg fully by setting \(q_a = 0\).
        For these simulations, we assume this can be done instantaneously.
\end{itemize}
While this method is titled ``rotation regulation", it is entirely possible to
choose \(p_\text{des} < \sqrt{60m^2gl^3}\);
this would stabilize an oscillation whose momentum at the bottom of the swing
arc is \(p_\text{des}\).

@@

The level set \(E_\pi\) with energy \(E(\pi,0)\) is outlined in black.
Recall that this level set is the boundary between oscillations and rotations of the
nominal pendulum.
The points where the orbit exits \(E_\pi\) are marked with black stars,
with the final departure marked by a red star.
Interestingly, our choice of \(I\) is large enough that we observe significant
differences between the nominal pendulum and the constrained dynamics:
\(E_\pi\) intersects the \(p_u\)-axis at \(\abs{p_u} \approx 0.18\), yet the
constrained acrobot begins rotating once it hits the
\(p_u\)-axis at \(\abs{p_u} \approx 0.16\). 
This indicates that higher values of \(I\) enable the acrobot to gain energy
faster and begin rotating sooner, so long as the actuator does not saturate.

@@

In simulation, we stabilized the dissipation VNHC and initialized the acrobot on
the constraint manifold with a rotation \((q_u,p_u) = (0,0.18)\).
We simulated the constrained system for \(30\) seconds and plotted the
resulting orbit in Figure \ref{fig:acrobot-diss-orbit}. 
As expected, the acrobot slows down over time.
We highlight the locations where the orbit crossed the set \(E_\pi\) by
black stars, with the final crossing in red.
After this final crossing, the acrobot ceased rotating and its oscillations
decayed to zero.

@@

Recall from Section \ref{sec:energy-reg} that one can use a supervisor to stabilize
oscillations by appropriately toggling between injection and dissipation VNHCs
whenever the orbit of the acrobot crosses the \(q_u\)-axis.
Figure \ref{fig:acrobot-osc-reg} shows the supervisor stabilizing an
oscillation with body angle \(q_\text{des} = \pi/2\) and a 
\(5\%\) hysteresis, meaning \(\delta = 0.05\).
The supervisor reevaluated its choice of VNHC at each black star;  
the red contour corresponds to the part of the orbit where the supervisor kept
the leg extended, because the oscillation was within tolerance of
\(q_\text{des}\).
The solid black line is the desired oscillation, and the dashed black
lines show the hysteresis around that orbit.

In Figure \ref{fig:acrobot-osc-reg-in} the acrobot was initialized at 
\((q_u,p_u) = (\pi/32,0)\); here the supervisor injected energy for most of the
orbit.
In Figure \ref{fig:acrobot-osc-reg-diss} the acrobot was initialized at the rotation
\((q_u,p_u) = (0,0.19)\); here the supervisor dissipated energy. 
In both cases, the supervisor stabilized the desired oscillation.

@@

\subsection{Rotation Regulation}
One can also use a supervisor to stabilize rotations through the mechanism
described in Section \ref{sec:energy-reg}, where the supervisor toggles between
injection and dissipation VNHCs at each crossing of the \(p_u\)-axis.
Rotation regulation for the acrobot is demonstrated in 
Figure \ref{fig:acrobot-rot-reg}, where the supervisor stabilizes 
\(p_\text{des} = 0.19\) with a \(2\%\) hysteresis \(\delta = 0.02\).
The supervisor evaluated its choice of VNHC at each black star.
Once it was within range of \(p_\text{des}\) it extended the legs completely,
the orbit of which is shown in red.

@@

Note the difference in shape between the blue rotations of the dissipation VNHC
and the red rotation of the nominal pendulum: the red one slows down much more
near \(|q_u| = \pi\).
This difference arises because of the size of \(I\):
if \(|I|\) were smaller, the blue rotations would be more similar in shape to the
red one because the constrained dynamics for the dissipation VNHC would be
well approximated by the nominal pendulum.

@@

We had hoped to show experimental results for dissipation and
oscillation/rotation regulation. 
Unfortunately the Dynamixel RX24F servo motor broke before we could perform
these experiments.
These motors are no longer in production at the time of writing this article, so
SUGAR has been retired from service.

@@

Here is fundamental idea behind this proof:
using a procedure devised by Mohammadi \etal
\cite{dynamic_vhcs_stabilize_closed_orbits}, on certain subsets 
\(D \subset \SxR\) one can find a diffeomorphism \(T(q_u,p_u) = (r,\theta)\)
into pseudo-polar coordinates adapted to level sets of \(E(q_u,p_u)\).
That is, when \(I = 0\) in \eqref{eqn:acrobot-constraint}, the pseudo-radius
\(r\) chooses a level set of \(E\) and the pseudo-angle \(\theta\) chooses a
point on that level set.
We can formally show that \(r\) increases ``on average", which means \(E\)
also increases on average and the constrained dynamics are gaining energy on
\(D\).
We formalize this idea in the following lemma.

@@

\begin{lemma}\label{lemma:energy-regulation-D}
    Consider the simplified acrobot \eqref{eqn:acrobot-hamiltonian} constrained
    by the VNHC \eqref{eqn:acrobot-constraint}, whose constrained dynamics are
    \eqref{eqn:acrobot-constrained-dynamics}.
    Let \(D \subset \SxR\) be open and 
    let \(]\underline{r}, \overline{r}[ \, \subset \R_{> 0}\).
    Suppose there exists a diffeomorphism 
    \begin{align*}
        T : D &\to ]\underline{r},\overline{r}[ \, \times \Sone
         \\
        (q_u,p_u) &\mapsto (r,\theta)
        ,
    \end{align*}
    into pseudo-polar coordinates adapted to level sets of \(E(q_u,p_u)\),
    satisfying \(\dot{r} \equiv 0\) and \(\dot{\theta} > 0\) when
    \(I = 0\).
    Let \(g(\theta,r,I) = \dot{r}/\dot{\theta}\) and assume that 
    \((\partial g/\partial r)(\theta,r,0) \equiv 0\).
    Assume \(b(\theta,r_0) = (\partial g/\partial I)(\theta,r_0,0)\)
    is \(2\pi\)-periodic in \(\theta\) for any fixed \(r_0\),
    and that there exists \(\epsilon > 0\) such that 
    \(S(r_0) := \int_0^{2\pi} b(\sigma,r_0)d\sigma \geq \epsilon\)
%    \[
%        S(r_0) := \int \limits_0^{2\pi} b(\sigma,r_0)d\sigma \geq \epsilon
%        ,
%    \]
    for all \(r_0 \in \, ]\underline{r},\overline{r}[\).
    Then there exists \(I^\star > 0\) small enough that the constrained dynamics
    gain energy on \(D\) for all \(I \in \,]0, I^\star]\), and lose energy on \(D\)
    for all \(I \in [-I^\star, 0[\).
\end{lemma}
\begin{proof}
    Since \(\dot{\theta} > 0\) when \(I = 0\) and the constrained dynamics 
    in \((r,\theta)\) coordinates are continuous in \(I\), there exists 
    \(I_1 > 0\) small enough that \(\dot{\theta} > 0\) for all
    \(I \in \, [-I_1,I_1]\).
    Hence, we can perform a time reparameterization \(t = t(\theta)\) to get a
    time-scaled pseudo-radius \(\hat{r}(\theta) := r(t(\theta))\).
    This reduces the system \((\dot{r},\dot{\theta})\) into the scalar
    time-varying ODE 
    \begin{equation}\label{eqn:rhat-ode}
        \begin{cases}
            \hat{r}^\prime= \left.\frac{\dot{r}}{\dot{\theta}}\right|_{r =
                \hat{r}(\theta)} = 
            g(\theta,\hat{r},I)
            , \\
            \hat{r}(0) = r_0
            ,
        \end{cases}
    \end{equation}
    where \(\theta\) now represents time,
    the prime denotes differentiation by \(\theta\), and
    \(r_0 \in \, ]\underline{r},\overline{r}[\) is the initial condition.

@@

We now apply the method described in
\cite{dynamic_vhcs_stabilize_closed_orbits} to find pseudo-polar coordinates
adapted to rotations.
These coordinates are 
\begin{align}
    \label{eqn:r-rot}
    r &= \sqrt{p_u^2 + 30m^2gl^3(1-c_u)}
    , \\
    \label{eqn:theta-rot}
    \theta &= \alpha(D) q_u
\end{align}
where \((r,\theta) \in \, ]\sqrt{60m^2gl^3}, \rho[ \times \Sone\),
and \(\alpha(D) = 1\) if \(D = \mathcal{R}^+(\rho)\)
or \(-1\) if \(D = \mathcal{R}^-(\rho)\).
The inverse of this transformation is
\begin{align*}
    q_u &= \alpha(D) \theta
    , \\
    p_u &= \alpha(D) \sqrt{r^2-30m^2gl^3(1-c_\theta)}
    .
\end{align*}

>>>>

% TODO: remove onecolumn and draftcls formatting
\documentclass[journal,twoside,web, twocolumn,draftcls]{ieeecolor}

@@

%---------- Header ----------% 
\newcommand*{\Title}{From gymnastics to virtual nonholonomic constraints: energy
injection, dissipation, and regulation for the acrobot}
\markboth{\journalname, VOL. XX, NO. XX, XXXX 2021}
{Moran-MacDonald \MakeLowercase{\textit{et al.}}: \Title}

@@

The definition above applies to vector fields that are a not necessarily
Hamiltonian, and therefore might not have an associated notion of ``energy". 
The terms ``gain energy" and ``lose energy" are to be interpreted loosely in
this context.
Our goal is to design a smooth function \(f : \Sone \times \R\) such that the
relation \(q_a = f(q_u,p_u)\) for system \eqref{eqn:acrobot-hamiltonian}
can be enforced asymptotically via feedback control (in Section \ref{sec:vnhc}
we call this a regular virtual nonholonomic constraint).  
We will further require that the dynamics of the acrobot, when the relation
holds, gain or lose energy on some set \(D\) in the sense of Definition
\ref{defn:energy-gain}. 
Note that any system satisfying Definition \ref{defn:energy-gain}
can have unstable equilibria on \(D\), but not limit cycles nor closed orbits.

%/========== VNHC ==========/%
\section{Theory of VNHCs}\label{sec:vnhc}
Before embarking on our design problem, we must summarize the relevant
theory of virtual nonholonomic constraints for a class of mechanical systems we
call ``simply actuated hamiltonian systems". 
The results we provide in Section \ref{sec:vnhc-vnhc} are not novel:
they are a special case of the results in
\cite{hybrid_zero_dynamics_bipedal_nhvcs}.

@@

\begin{assm}\label{assm:B-const}
    The input matrix \(B(q) \equiv B \in \R^{n\times k}\) is constant
    and has full rank \(k < n\).
\end{assm}

%\begin{assm}\label{assm:B-perp}
%    There exists a right semi-orthogonal matrix 
%    \(B^\perp \in \R^{(n-k)\times n}\)
%    which is a left-annihilator for \(B\). 
%\end{assm}
%
%Note that Assumption \ref{assm:B-perp} requires the rows of \(B^\perp\) be unit vectors
%that are mutually orthogonal. 
%When \(k = (n-1)\), Assumption \ref{assm:B-perp} can be removed because it is
%automatically implied by Assumption \ref{assm:B-const}.
%
When \(\mathcal{Q} = \R^n\), the above assumption allows us to define a
canonical coordinate transformation of \eqref{eqn:hamiltonian} 
which decouples the input forces.
To define this transformation we will make use of the following lemma.

@@

In light of Lemma \ref{lemma:B-orthogonal} there is no loss of generality in
assuming that the input matrix is left semi-orthogonal, which means that
\(B\tpose\) is right semi-orthogonal.
Now, let \(\mathbf{B} := [B^\perp; B\tpose]\) where 
\(B^\perp \in \R^{(n-k)\times k}\) is a left annihilator of \(B\), i.e.,
\(B^\perp B = \Zmat{(n-k) \times k}\).
Since \(B\) is constant, such a \(B^\perp\) exists and
\(\mathbf{B}\) is invertible.

The following theorem requires that \(\mathcal{Q} = \R^n\) so that the
transformation \(\tilde{q} = \mathbf{B}q\) is well defined.

\begin{thm}\label{thm:simply-actuated}
    Take the Hamiltonian system \eqref{eqn:hamiltonian} with configuration
    manifold \(\mathcal{Q} = \R^n\) and suppose
    Assumption \ref{assm:B-const} holds.
    The coordinate transformation
    \(\left(\tilde{q} = (\mathbf{B}\tpose)\inv q, \tilde{p} = \mathbf{B}p\right)\)
    is a canonical transformation and the resulting dynamics are given by 
    \begin{gather}\label{eqn:simple-hamiltonian}
        \mathcal{H}(\tilde{q},\tilde{p}) = 
        \frac{1}{2} \tilde{p}\tpose \Minv(\tilde{q}) \tilde{p} + V(\tilde{q})
        , \\
       \begin{cases}
           \dot{\tilde{q}} = \Minv(\tilde{q})\tilde{p}
           , \\
           \dot{\tilde{p}} = -\frac{1}{2} (\Id{n} \otimes \tilde{p}\tpose)
           \nabla_{\tilde{q}} \Minv(\tilde{q}) \tilde{p} \\
           \phantom{---} - \nabla_{\tilde{q}} V(\tilde{q}) + \simpleB \tau
            ,
        \end{cases} \nonumber
    \end{gather}
    where 
    \(\Minv(\tilde{q}) := 
    (\mathbf{B}\tpose)\inv D^{-1}(\mathbf{B}\tpose \tilde{q})\mathbf{B}\inv\)
    and
    \(V(\tilde{q}) := P(\mathbf{B}\tpose \tilde{q})\).
\end{thm}
\begin{proof}
    Since \(\mathbf{B}\) is constant, this transformation satisfies
    \(\partial\tilde{q}_i/\partial p_j = \partial\tilde{p}_i/\partial q_j = 0\) for all 
    \(i,j \in \{1,\ldots,n\}\).
    This implies the Poisson brackets \([\tilde{q}_i, \tilde{q}_j]\)
    and \([\tilde{p}_i,\tilde{p}_j]\) are both zero.
    Then, one can show that
    \([\tilde{p}_i, \tilde{q}_j] = (\mathbf{B}\tpose)\inv_i (\mathbf{B}\tpose)_j
        = \delta_i^j\).
    By (45.10) in \cite{landau_mechanics}, this is a canonical transformation
    and the new Hamiltonian is
    \(\mathcal{H}(\mathbf{B}\tpose \tilde{q}, \mathbf{B}\inv \tilde{p})\).
    Finally, since \(\dot{\tilde{p}} = \mathbf{B} \dot{p}\), the input
    matrix for the system in \((\tilde{q},\tilde{p})\) coordinates is
    \(\mathbf{B}B = [\Zmat{(n-k)\times k}; \Id{k}]\), which proves the theorem.
\end{proof}

We call the \((\tilde{q},\tilde{p})\) coordinates
\textit{simply actuated coordinates}, and we call any Hamiltonian system
whose input matrix is \([\Zmat{(n-k)\times k}; \Id{k}]\) a 
\textit{simply actuated hamiltonian system}.
The first \((n-k)\) configuration variables in \(\tilde{q}\), labelled \(q_u\),
are the \textit{unactuated coordinates}; 
the remaining \(k\) configuration variables, labelled \(q_a\), are the
\textit{actuated coordinates}.
The corresponding \((p_u, p_a)\) in \(\tilde{p}\) are the \textit{unactuated}
and \textit{actuated momenta}, respectively.

In this article we focus on the acrobot which, despite having a configuration
manifold which is not \(\R^n\), is already a simply actuated Hamiltonian system.

@@

\subsection{Energy Regulation}\label{sec:energy-reg}
One can apply the results of Theorem \ref{thm:acrobot-energy-stabilization} 
towards energy regulation; 
that is, one can stabilize oscillations or rotations by appropriately toggling
between injection and dissipation VNHCs, which can be achieved by changing the
sign of \(I\) in \eqref{eqn:acrobot-constraint}.
The acrobot is \textit{oscillating} when it swings back and forth without
performing a full revolution around the pivot point.
Oscillations correspond to \(p_u\) changing sign periodically, and the peak
torso angle (i.e., the point where an orbit crosses the \(q_u\)-axis)
is called the \textit{oscillation angle}.
Conversely, the acrobot is \textit{rotating} when its torso is performing full
revolutions around the pivot, which corresponds mathematically to \(p_u\)
maintaining the same sign.
The momentum achieved by the acrobot at the bottom of the swing
(i.e., the point where an orbit of rotation crosses the \(p_u\)-axis) is
called the \textit{rotation rate}.

\textbf{Rotation Regulation}: one chooses a desired rotation rate 
\(p_\text{des} > 0\) and a hysteresis value \(\delta \in [0,1]\).
Each time the orbit crosses the \(p_u\)-axis (i.e. when \(q_u = 0\)), the
supervisor changes which VNHC is enforced as follows:
\begin{itemize}
    \item If \(|p_u| < (1-\delta)p_\text{des}\), enable the injection VNHC by
        setting \(I > 0\) in \eqref{eqn:acrobot-constraint}.
    \item If \(|p_u| > (1+\delta)p_\text{des}\), enable the dissipation VNHC by
        replacing \(I > 0\) with \(-I < 0\) in \eqref{eqn:acrobot-constraint}.
    \item If \((1-\delta)p_\text{des} \leq |p_u| \leq (1+\delta)p_\text{des}\),
        extend the leg fully by setting \(q_a = 0\).
        For these simulations, we assume this can be done instantaneously,
        though in practice this is impossible.
\end{itemize}
Note that all orbits of the acrobot must cross the \(p_u\) axis, 
so this method will stabilize a rotation even if the acrobot is initialized in
\(\mathcal{O}_1\).
Also, the above supervisor is designed to regulate a rotation rate, but it
cannot regulate the rotation direction.

The term ``rotation regulation" implies that one must choose
\(p_\text{des}\) and \(\delta\) so that 
\((1-\delta) p_\text{des} > \sqrt{60m^2gl^3}\).
In fact, it is entirely possible to choose \(p_\text{des}\) and \(\delta\) so
that \((1+\delta)p_\text{des} \leq \sqrt{60m^2gl^3}\).
Rather than stabilizing a rotation, the supervisor would then stabilize an
oscillation whose momentum at the bottom of the swing arc is \(p_\text{des}\).
However, there is a preferred method for stabilizing oscillations
which we outline next.

\textbf{Oscillation Regulation}: one chooses a desired oscillation angle 
\(q_\text{des} \in \, ]0,\pi[\) and, to avoid infinite switching, a
hysteresis value \(\delta \in [0,\pi - q_\text{des}]\).
An orbit in the \((q_u,p_u)\)-plane will either cross the \(q_u\) axis if the
orbit corresponds to a rocking motion, or it will cross the line \(|q_u| = \pi\)
if the orbit corresponds to a full revolution.
When either of these occur, the supervisor does the following: 
\begin{itemize}
    \item If \(|q_u| < (1-\delta)q_\text{des}\), enable the injection VNHC.
    \item If \(|q_u| > (1+\delta)q_\text{des}\), enable the dissipation VNHC.
    \item If \((1-\delta)q_\text{des} \leq |q_u| \leq (1+\delta)q_\text{des}\),
        keep the leg extended at \(q_a = 0\). This can be done continously since
        \(q_a = 0\) when \(p_u = 0\).
\end{itemize}
Note that by our choice of \(\delta\), if the supervisor kicks in when 
\(|q_u| = \pi\) (i.e., when the robot is rotating) then the supervisor will
automatically enable the dissipation VNHC.

@@

The level set \(E_\pi\) with energy \(E(\pi,0)\) is outlined in black.
Recall that this level set is the boundary between oscillations and rotations of the
nominal pendulum.
The points where the orbit exits \(E_\pi\) are marked with black asterisks,
with the final departure marked by a red asterisk.
Interestingly, our choice of \(I\) is large enough that we observe significant
differences between the nominal pendulum and the constrained dynamics:
\(E_\pi\) intersects the \(p_u\)-axis at \(\abs{p_u} \approx 0.18\), yet the
constrained acrobot begins rotating once it hits the
\(p_u\)-axis at \(\abs{p_u} \approx 0.16\). 
This indicates that higher values of \(I\) enable the acrobot to gain energy
faster and begin rotating sooner, so long as the actuator does not saturate.

@@

In simulation, we stabilized the dissipation VNHC and initialized the acrobot on
the constraint manifold with a rotation \((q_u,p_u) = (0,0.18)\).
We simulated the constrained system for \(30\) seconds and plotted the
resulting orbit in Figure \ref{fig:acrobot-diss-orbit}. 
As expected, the acrobot slows down over time.
We highlight the locations where the orbit crossed the set \(E_\pi\) by
black asterisks, with the final crossing in red.
After this final crossing, the acrobot ceased rotating and its oscillations
decayed to zero.

@@

Recall from Section \ref{sec:energy-reg} that one can use a supervisor to stabilize
oscillations by appropriately toggling between injection and dissipation VNHCs
whenever the orbit of the acrobot crosses the \(q_u\)-axis.
Figure \ref{fig:acrobot-osc-reg} shows the supervisor stabilizing an
oscillation with body angle \(q_\text{des} = \pi/2\) and a 
\(5\%\) hysteresis, meaning \(\delta = 0.05\).
The supervisor reevaluated its choice of VNHC at each black asterisk;  
the red contour corresponds to the part of the orbit where the supervisor kept
the leg extended, because the oscillation was within tolerance of
\(q_\text{des}\).
The solid black line is the desired oscillation, and the dashed black
lines show the hysteresis around that orbit.

In Figure \ref{fig:acrobot-osc-reg-in} the acrobot was initialized at 
\((q_u,p_u) = (\pi/32,0)\); here the supervisor injected energy to stabilize the
desired orbit.
In Figure \ref{fig:acrobot-osc-reg-diss} the acrobot was initialized at the rotation
\((q_u,p_u) = (0,0.19)\); note that the supervisor is first enabled at the line
\(|q_u| = \pi\) and dissipates energy. Eventually the acrobot begins a rocking
motion and the supervisor is enable again each time the acrobot hits the
\(q_u\)-axis. 
The supervisor continues to dissipate energy until it stabilizes the desired
oscillation.

@@

\subsection{Rotation Regulation}
One can also use a supervisor to stabilize rotations through the mechanism
described in Section \ref{sec:energy-reg}, where the supervisor toggles between
injection and dissipation VNHCs at each crossing of the \(p_u\)-axis.
Rotation regulation for the acrobot is demonstrated in 
Figure \ref{fig:acrobot-rot-reg}, where the supervisor stabilizes 
\(p_\text{des} = 0.19\) with a \(2\%\) hysteresis \(\delta = 0.02\).
The supervisor evaluated its choice of VNHC at each black asterisk.
Once it was within range of \(p_\text{des}\) it extended the legs completely,
the orbit of which is shown in red.

@@

Note the difference in shape between the blue rotations of the dissipation VNHC
and the red rotation of the nominal pendulum in Figure
\ref{fig:acrobot-rot-reg}: the red one slows down much more
near \(|q_u| = \pi\).
This difference arises because of the size of \(I\):
if \(|I|\) were smaller, the blue rotations would be more similar in shape to the
red one because the constrained dynamics for the dissipation VNHC would be
well approximated by the nominal pendulum.

@@

@@

Here is fundamental idea behind this proof:
By means of a diffeomorphism \(T(q_u,p_u) = (r,\theta)\) we transform the
constrained dynamics \eqref{eqn:acrobot-constrained-dynamics} 
into pseudo-polar coordinates adapted to level sets of \(E(q_u,p_u)\).
That is, when \(I = 0\) in \eqref{eqn:acrobot-constraint}, the pseudo-radius
\(r\) chooses a level set of \(E\) and the pseudo-angle \(\theta\) chooses a
point on that level set.
We can formally show that \(r\) increases ``on average", which means \(E\)
also increases on average and the constrained dynamics are gaining energy on
\(D\).
We formalize this idea in the following lemma.

@@

\begin{lemma}\label{lemma:energy-regulation-D}
    Consider the simplified acrobot \eqref{eqn:acrobot-hamiltonian} constrained
    by the VNHC \eqref{eqn:acrobot-constraint}, whose constrained dynamics are
    \eqref{eqn:acrobot-constrained-dynamics}.
    Let \(D \subset \SxR\) be open and 
    let \(]\underline{r}, \overline{r}[ \, \subset \R_{> 0}\).
    Suppose there exists a diffeomorphism 
    \begin{align*}
        T : D &\to ]\underline{r},\overline{r}[ \, \times \Sone
         \\
        (q_u,p_u) &\mapsto (r,\theta)
        ,
    \end{align*}
    satisfying \(\dot{r} \equiv 0\) and \(\dot{\theta} > 0\) when
    \(I = 0\).
    Let \(g(\theta,r,I) = \dot{r}/\dot{\theta}\) and assume that 
    \((\partial g/\partial r)(\theta,r,0) \equiv 0\).
    Assume \(b(\theta,r_0) = (\partial g/\partial I)(\theta,r_0,0)\)
    is \(2\pi\)-periodic in \(\theta\) for any fixed \(r_0\),
    and that there exists \(\epsilon > 0\) such that 
    \(S(r_0) := \int_0^{2\pi} b(\sigma,r_0)d\sigma \geq \epsilon\)
%    \[
%        S(r_0) := \int \limits_0^{2\pi} b(\sigma,r_0)d\sigma \geq \epsilon
%        ,
%    \]
    for all \(r_0 \in \, ]\underline{r},\overline{r}[\).
    Then there exists \(I^\star > 0\) small enough that the constrained dynamics
    gain energy on \(D\) for all \(I \in \,]0, I^\star]\), and lose energy on \(D\)
    for all \(I \in [-I^\star, 0[\).
\end{lemma}
\begin{proof}
    Since \(\dot{\theta} > 0\) when \(I = 0\) and the constrained dynamics 
    in \((r,\theta)\) coordinates are continuous in \(I\), there exists 
    \(I_1 > 0\) small enough that \(\dot{\theta} > 0\) for all
    \(I \in \, [-I_1,I_1]\).
    Hence, we can perform a time reparameterization \(t = t(\theta)\) to get a
    time-scaled pseudo-radius \(\hat{r}(\theta) := r(t(\theta))\).
    This reduces the system \((\dot{r},\dot{\theta})\) into the scalar
    time-varying ODE 
    \begin{equation}\label{eqn:rhat-ode}
        \begin{cases}
            \hat{r}^\prime= \left.\frac{\dot{r}}{\dot{\theta}}\right|_{r =
                \hat{r}(\theta)} = 
            g(\theta,\hat{r},I)
            , \\
            \hat{r}(0) = r_0
            ,
        \end{cases}
    \end{equation}
    where \(\theta\) now represents time,
    the prime denotes differentiation by \(\theta\), and
    \(r_0 \in \, ]\underline{r},\overline{r}[\) is the initial condition.

@@

We now apply
\cite[Eqn. (8)]{dynamic_vhcs_stabilize_closed_orbits} to find pseudo-polar coordinates
adapted to rotations.
These coordinates are 
\begin{align}
    \label{eqn:r-rot}
    r &= \sqrt{p_u^2 + 30m^2gl^3(1-c_u)}
    , \\
    \label{eqn:theta-rot}
    \theta &= \alpha(D) q_u
\end{align}
where \((r,\theta) \in \, ]\sqrt{60m^2gl^3}, \rho[ \times \Sone\),
and \(\alpha(D) = 1\) if \(D = \mathcal{R}^+(\rho)\)
or \(-1\) if \(D = \mathcal{R}^-(\rho)\).
The inverse of this transformation is
\begin{align*}
    q_u &= \alpha(D) \theta
    , \\
    p_u &= \alpha(D) \sqrt{r^2-30m^2gl^3(1-c_\theta)}
    .
\end{align*}
